Optogenetic Maxwell Demon to Exploit Intrinsic Noise and Control Cell Differentiation Despite Time Delays and Extrinsic Variability

### 2023-02-01 paper.tex
More updates to introduction.

Classical and modern control methods like PID control and model predictive control have been implemented in such systems\cite{} which can control synthetic systems to different stable points. Theoretical works have shown that stochastic systems can be controlled to unstable points by oscillating around those points \cite{Guarino2020} but new control techniques which exploit the full probability distribution information of the system can be used in certain scenarios to control simple systems to unstable points by exploiting the natural noise of protein production to its advantage. Previous control effort focus on deterministic ODE formulations of the control problem, but only a few studies seek to analyze the full chemical master equation to exploit noise to {\em improve} the control of synthetic biology processes \cite{Szymanska2015,May2021}.

Szymanska et al.~\cite{Szymanska2015} examined a theoretical system inspired by the genetic toggle switch from Kobayashi et al.\cite{Kobayashi:2004}, where two gene products repress one another, but where degradation of one protein was under the control of UV radiation via the XXX pathway. Their analysis showed that control was possible despite certain parameter errors and time delays due to maturation of fluorescent proteins and limited observation of the regulatory proteins. 
To explore this concept in the context of more realistic parameters, May et al.~\cite{May2021} fit a simplified stochastic model to reproduce data measured in Baumschlager, et al.\cite{} for the expression from the XXX promoter under the optogenetic control of a UV-activated T7 polymerase. An extended model showed that addition of a positive auto-regulation module (see Figure XXX) could help to maintain an elevated expression phenotype in the presence of UV excitation. Using discrete stochastic models based on direct solutions to the chemical master equation, it was shown that   the combination of biochemical noise, nonlinear auto-regulation, and a {\em single} optogenetic feedback could allow two genetically identical cells with arbitrary initial conditions to be controlled simultaneously to reach different specified phenotypes (Figure \ref{cartoons}A). Moreover, effective control was possible not only for a Fully Aware Feedback Controller (FAC) that observed the protein counts of all observed cells, but control could be achieved almost as well for a Partially Aware Feedback Controller (PAC) that observed the protein counts of only one cell and used a probabilistic model to estimate the expression for unobserved cells. Moreover, it was again seen that control remained effective even in the circumstance where the model used to design the control law was an inexact simplification of the process dynamics to which the controller was later applied.  
%The solutions of the controllers was carried out by solving the chemical master equation (i.e., the forward Komogorov equation) to estimate the probability of possible cellular states. For a two cell, single-input-single output (SIMO) control problem, the FAC performed the best (figure\ \ref{cartoons}C(top)), and the PAC (figure\ \ref{cartoons}C(bottom)) 2nd, and pMPC controllers 3rd. In all cases, the controller's knowledge and explicit treatment of intrinsic noise was critical to improve the control performance, break symmetry, and enable the SIMO control of all cells.

In this paper, the control problem is extended to examine how the fully- and partially-aware controllers (FAC and PAC, respectively) are affected by additional model uncertainties, including coarse grained approximations to the system dynamics, uncertainties or errors in system parameters, and delays in the biochemical sensors or optogenetic controllers through which the external feedback is implemented. First, in the next section, we briefly describe our methods to set up a stochastic chemical master equation model for multiple individual cells in the same spatially homogeneous environment; we formulate the FAC and PAC controllers on the dynamics of cells within that environment; and we explain how the control law is optimized to achieve desired performance objectives. Next, in the `Results' section, we examine how the granularity of the model formulation, the inaccuracy of parametric assumptions, and the inclusion of time delays affect the FAC and PAC control performance. Finally, in `Conclusions' we summarize our findings and discuss the implications that these controllers, and their robustness to modeling errors, may have on future systems in synthetic biology.

>>>>

Classical and modern control methods like PID control and model predictive control have been implemented in such systems\cite{} which can control synthetic systems to different stable points. 
% XXX Missing Citation above.
Theoretical works have shown that stochastic systems can be controlled to unstable points by oscillating around those points \cite{Guarino2020} but new control techniques which exploit the full probability distribution information of the system can be used in certain scenarios to control simple systems to unstable points by exploiting the natural noise of protein production to its advantage. Previous control effort focus on deterministic ODE formulations of the control problem, but only a few studies seek to analyze the full chemical master equation to exploit noise to {\em improve} the control of synthetic biology processes \cite{Szymanska2015,May2021}.

Szymanska et al.~\cite{Szymanska2015} examined a theoretical system inspired by the genetic toggle switch from Kobayashi et al.\cite{Kobayashi:2004}, where two gene products repress one another, but where degradation of one protein was under the control of UV radiation via the XXX pathway. 
% XXX - add citation to Kobayashi paper, and update name of the pathway.  I think it was the SOS pathway.
Their analysis showed that control was possible despite certain parameter errors and time delays due to maturation of fluorescent proteins and limited observation of the regulatory proteins. 
To explore this concept in the context of more realistic models and parameters, May et al.~\cite{May2021} fit a simplified stochastic model to reproduce data measured in Baumschlager, et al.\cite{} for the expression from the XXX promoter under the optogenetic control of a UV-activated T7 polymerase (see model in Figure\ \ref{cartoons}A, top promoter). 
% XXX - add details about the promoter and citation.
% XXX - let's change the order of panels A and B for this figure so that we can refer to them in the proper order.  I adjusted the text below to assume this change.
An extended model showed that addition of a positive auto-regulation module (Figure\ \ref{cartoons}A, bottom promoter) could help to maintain an elevated expression phenotype in the presence of UV excitation.
% XXX - Missing figure reference
Using discrete stochastic models based on direct solutions to the chemical master equation, it was shown that the combination of biochemical noise, nonlinear auto-regulation, and a {\em single} optogenetic feedback could allow two genetically identical cells with arbitrary initial conditions to be controlled simultaneously to reach different specified phenotypes (Figure \ref{cartoons}B). 
For example, Figure\ \ref{cartoons}C(left) reproduces the heat maps for two control laws, where the level of a single external UV excitation signal was modulated depending on the observed cellular activity, and Figure\ \ref{cartoons}C(right) shows the corresponding marginal stationary distributions that could be achieved under these control laws. 
Effective control was possible not only for a Fully Aware Feedback Controller (FAC, Figure\ \ref{cartoons}C(top)) that observed the protein counts of all observed cells, but control could be achieved nearly as well for a Partially Aware Feedback Controller (PAC, Figure\ \ref{cartoons}C(bottom)) that observed the protein counts of only one cell. 
Moreover, it was again seen that these control laws remained effective even in the circumstance where the model used to design the control law was an inexact simplification of the process dynamics to which the controller was later applied.  
%The solutions of the controllers was carried out by solving the chemical master equation (i.e., the forward Komogorov equation) to estimate the probability of possible cellular states. For a two cell, single-input-single output (SIMO) control problem, the FAC performed the best (figure\ \ref{cartoons}C(top)), and the PAC (figure\ \ref{cartoons}C(bottom)) 2nd, and pMPC controllers 3rd. In all cases, the controller's knowledge and explicit treatment of intrinsic noise was critical to improve the control performance, break symmetry, and enable the SIMO control of all cells.

In this paper, the analysis of this Single-Input-Multiple-Output (SIMO) multicellular control problem is extended to examine how the FAC and PAC controllers are affected by additional model uncertainties, including coarse grained approximations to the system dynamics, uncertainties or errors in system parameters, and delays in the biochemical sensors or optogenetic controllers through which the external feedback is implemented. In the next section, we briefly describe our methods to formulate the chemical master equation to analyze multiple individual cells in the same spatially homogeneous environment; we define the FAC and PAC controllers and their effects the dynamics of cells within that environment; and we explain how the control law is optimized to achieve desired performance objectives. Next, in the `Results' section, we explore how the granularity of the model formulation, the inaccuracy of parametric assumptions, and the inclusion of time delays would affect the FAC and PAC control performance. Finally, in `Conclusions' we summarize our findings and discuss the implications that these controllers, and their robustness to modeling errors, may have on future systems in synthetic biology.
